---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Travis-CI Build Status](https://travis-ci.org/mdsumner/quadmesh.svg?branch=master)](https://travis-ci.org/mdsumner/quadmesh)
[![CRAN status](https://www.r-pkg.org/badges/version/quadmesh)](https://cran.r-project.org/package=quadmesh)


# quadmesh


A *quadmesh* is a dense mesh describing a topologically continuous surface of 4-corner primitives. This is also known as a *cell-based raster* but in those contexts the corner coordinates and the continuous nature of the mesh is completely implicit. 

This package provides helpers for working with mesh interpretation of gridded data to enable

* arbitrary reprojection of raster cells without information loss.
* easy plotting of grids in 3D visualization tools (like `rgl`).
* the corner-based interpretation of a grid.
* fast polygonization of individual cells. 

You can install:

* the latest released version from CRAN with

```{r,eval=FALSE}
install.packages("quadmesh")
````

* the latest development version from Github with

```{r,eval=FALSE}
devtools::install_github("mdsumner/quadmesh")
```



Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.
