---
title: "Quadmesh"
author: "Michael D. Sumner"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{quadmesh}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# quadmesh

Build a "quadmesh"" in R. 

```{r}
library(quadmesh)
library(raster)

data(volcano)
r <- setExtent(raster(volcano), extent(0, 100, 0, 200))


qm <- quadmesh(r)

library(rgl)
scl <- function(x) (x - min(x))/diff(range(x))
shade3d(qm, col = grey(scl(qm$vb[3,qm$ib])))

rglwidget()
```

A "quadmesh" is a dense mesh describing a topologically continuous surface of 4-corner primitives. I.e. a grid, without the "regular". This is useful particularly when combined with map projections and texture mapping. 

We are not limited to a regular grid, trivially let's distort the mesh by a weird scaling factor. 

The topology of the grid is still sound, but we are no longer bound to the regular constraint. 

```{r}
qm <- quadmesh(r)

qm$vb[1,] <- qm$vb[1,] * qm$vb[2,]/54
open3d()
shade3d(qm, col = grey(scl(qm$vb[3,qm$ib])))

rglwidget()
```


